---
title: "HCMR Cross-Trait Association Analysis"
author: "Jonathan Chan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)
library(ggsignif)
library(car)
rm(list=ls())

#Setting theme for ggplot to classic - You can change your plot text font and size here
theme_set(
  theme_classic()
  + theme(text = element_text(family = 'sans', size=14))
  ) 
```

This notebook looks at cross-trait associations within HCMR, evaluating them using multivariate linear regression of both joint and marginal models to correct for confounders such as age;sex; BMI and BSA (where necessary).

# Import

```{r import}
#This is as per OMGL classifications
hcmr_rarevar <- read_tsv('DATA/HCMR/hcmr_rarevar_class_plp_vusfpvus.tsv') %>%
  dplyr::rename('rarevar_class'=class, 'HCR_IDs'='HCR_ID')

total_df_notransform <- readRDS('DATA/HCMR/total_notransform_df.rds') %>%
  mutate(ecvfwhole=ecvfwhole*100)

total_i_all_cont_phenotypes <- readRDS('DATA/HCMR/total_i_all_cont_phenotypes.rds')

total_df_notransform_rare <- left_join(total_df_notransform, select(hcmr_rarevar, HCR_IDs, rarevar_class))
total_i_all_cont_phenotypes <- c(total_i_all_cont_phenotypes, ncol(total_df_notransform_rare)-1,ncol(total_df_notransform_rare))
rm(total_df_notransform)

label_tb <- read_tsv('DATA/HCMR/combo_labels.tsv') 
label_tb$Label[label_tb$Variable=='ecvfwhole'] <- 'Whole LV ECVF (based on 3 slices) (%)'


#Subset to variables of interest
i_cont_biomarker <- seq(139,148)
vars_of_interest <- c('HCR_IDs','age','gender','height','weight','bmi','bsa','blnyha','cmrsysbp','cmrdiabp', colnames(total_df_notransform_rare)[i_cont_biomarker], 'lge_total','lvmassi','ci','wallthkmax','lvef','ecvfwhole','lvs_eccmean','lvs_ellmean','lvs_errmean','rarevar_class','sarcomere','lge_pres')

#Select all of the variables of interest
filtered_tb <- total_df_notransform_rare %>% 
  select(all_of(vars_of_interest)) 


#Import in the 10PCs generated in HCMR only
eigenvectors <- read_tsv('DATA/HCMR/eigenvectors_hcmr_10pcs') %>%
  rename_with(~str_replace(.,'U','pc'),contains('U'))

filtered_tb <- filtered_tb %>%
  left_join(select(eigenvectors, IID, pc1:pc5), by=c('HCR_IDs' = 'IID')) %>%
  filter(!is.na(pc1), !is.na(age), !is.na(gender), !is.na(bmi))

print(str_c(nrow(filtered_tb), ' HCMR patients are analysed with non-NA values in age; gender; BMI and common genetic PCs'))
```
# Normalisation

For ease of comparison, I will transform all the biomarkers using inverse-rank based transformation.

For comparison sakes (i.e to compare across the phenotypes individually), I also do the same across all CMR endophenotypes.

```{r normalisation}
log_transformer <- function(data,columns){ #Columns are a vector of characters
  
  for (i in columns){
    data[[i]] <- log1p(data[[i]])
    data[[i]] <- scale(data[[i]])
    hist(data[[i]], main=i,xlab=i)
  }
  
  return(data)
}

inverse_rank_transformer <- function(data,columns){ #Columns are a vector of characters
  
  for (i in columns){
    col_label <- i
    col_label_before <- columns[match(i,columns)-1]
    mask <- !is.na(data[[i]]) #Mask to filter out NA values
                             
    # Filter out NA values for that phenotype
    data_copy <- subset(data, mask)
    data_copy[[i]] <- RNOmni::RankNorm(data_copy[[i]]) #Performing rank-inverse normalisation
    data_copy <- select(data_copy, 'HCR_IDs',all_of(i)) #Taking only the column with SitePatID and the column of interest
    
    data <- select(data, - i) %>% #Removing the original column and adding back the new column after rank-inverse normalisation
      left_join(data_copy, by='HCR_IDs') %>%
      relocate(i, .after=!!col_label_before) #Moving the appended column back to its original index
    
    hist(data[[i]], main=i,xlab=i)
  }
  return(data)
}

# log_transform_columns <- c(
#   vars_of_interest[11:20]
# )
# log_transform_columns <- log_transform_columns[!log_transform_columns %in% c('TnTStat','mmp1_timp1','NTproBNP','gal3','st2', 'mmp1','timp1','cicp','bap','cicp_bap')]
# log_transformed_tb <- log_transformer(filtered_tb, log_transform_columns)
```


```{r normalisation}
#Define the groups of variables used
pp <- c('NTproBNP','TnTStat','gal3','st2','mmp1','cicp')
fibr <- c('lge_total','ecvfwhole')
func <- c('lvef','lvs_ellmean','lvs_eccmean','lvs_errmean','ci')
hty <- c('lvmassi')
```


```{r normalisation}
#Inverse based rank transform pp
rank_transform_columns <- c(pp,fibr,func, hty, 'wallthkmax')
# rank_transform_columns <- c(pp)
rank_transformed_tb <- inverse_rank_transformer(filtered_tb, rank_transform_columns)

#Merge back everything together

transformed_tb <- select(filtered_tb, colnames(filtered_tb)[!colnames(filtered_tb) %in% c(#log_transform_columns, 
  rank_transform_columns)]) %>%
  #bind_cols(select(log_transformed_tb, log_transform_columns)) %>%
  bind_cols(select(rank_transformed_tb, rank_transform_columns)) 
  # %>% mutate(cv_prs=scale(cv_prs)) #Also need to scale CV_PRS

rm(log_transformed_tb, rank_transformed_tb)

#Adjust the label tb to represent the transformations
label_tb <- label_tb %>% 
  mutate(Label = case_when(Variable %in% c(#log_transform_columns, 
    rank_transform_columns) ~ str_c(str_match(Label,'(^[^\\(]+)')[,2],
                                                                                                  ' (SD)'),
                           T~Label))
  

```
### Summary Table

This outputs a summary table for publication.

```{r}
#Only rank-transform the pp and not the CMR parameters

#Inverse based rank transform pp
rank_transform_columns <- c(pp)
rank_transformed_tb <- inverse_rank_transformer(filtered_tb, rank_transform_columns)

#Merge back everything together

pponly_transformed_tb <- select(filtered_tb, colnames(filtered_tb)[!colnames(filtered_tb) %in% c(#log_transform_columns, 
  rank_transform_columns)]) %>%
  #bind_cols(select(log_transformed_tb, log_transform_columns)) %>%
  bind_cols(select(rank_transformed_tb, rank_transform_columns)) 
  # %>%mutate(cv_prs=scale(cv_prs)) #Also need to scale CV_PRS

rm(log_transformed_tb, rank_transformed_tb)
```

```{r}
covars <- c('age','sex','self_reported_ethnicity','bmi','bsa','cmrdiabp','sarcomere')

#Run pairwise Spearman rank correlation matrix between each of the continuous covariates (not sex, ethnicity or sarcomere) in the pponly_transformed_tb
quant_covars <- covars[!covars %in% c('sex','self_reported_ethnicity','sarcomere')]
cor_matrix <- pponly_transformed_tb %>%
  select(all_of(quant_covars)) %>%
  cor(method='spearman', use='pairwise.complete.obs')
print(cor_matrix)

highly_correlated_pairs <- which(abs(cor_matrix) > 0.8 & abs(cor_matrix) < 1.00, arr.ind = TRUE)
for (i in seq_len(nrow(highly_correlated_pairs))) {
  row <- highly_correlated_pairs[i, 1]
  col <- highly_correlated_pairs[i, 2]
  cat(sprintf("Highly correlated pair: %s and %s with correlation %.2f\n",
              colnames(cor_matrix)[row],
              colnames(cor_matrix)[col],
              cor_matrix[row, col]))
}

#Add back the self-reported ethnicity to 
summary_tb <- pponly_transformed_tb %>%
  left_join(select(total_df_notransform_rare,HCR_IDs,race)) %>%
  select(-starts_with('pc')) %>%
  dplyr::rename('sex'='gender','self_reported_ethnicity'='race') %>%
  select(all_of(c(covars,'wallthkmax',hty, func,fibr,pp)))

#Evaluate normality of each continuous variable
walk2(select(summary_tb, all_of(pp)), pp,~print(hist(.x, xlab = .y)))
walk2(select(summary_tb, all_of(covars[!covars %in% c('sex','self_reported_ethnicity','sarcomere')])), covars[!covars %in% c('sex','self_reported_ethnicity','sarcomere')],~print(hist(.x, xlab = .y)))
walk2(select(summary_tb, all_of(c('wallthkmax',hty,func,fibr))), c('wallthkmax',hty,func,fibr),~print(hist(.x, xlab = .y)))
```

```{r}

```
```{r}
generate_summary_table_R <- function(input_df,
                                     output_tsv_path,
                                     mean_sd_vars = character(0),
                                     median_iqr_vars = character(0)) {
  # Check if input is a data frame
  if (!is.data.frame(input_df)) {
    stop("Input must be a data frame or tibble.")
  }

  # Get column names from the input data frame
  col_names <- names(input_df)

  # Handle the case of an empty data frame (0 columns)
  if (length(col_names) == 0) {
    summary_tibble <- tibble(
      Original_Column = character(),
      Summary = character(),
      Non_NA_Count = integer()
    )
  } else {
    # Use purrr::map_dfr to iterate over column names and create a summary row for each
    summary_tibble <- purrr::map_dfr(col_names, function(col_name) {
      column_vector <- dplyr::pull(input_df, col_name)
      non_na_count <- sum(!is.na(column_vector))
      summary_str <- "Error: Type not handled" # Default if no specific logic matches

      # 1. Qualitative columns (character, factor, logical)
      if (is.character(column_vector) || is.factor(column_vector) || is.logical(column_vector)) {
        if (non_na_count > 0) {
          temp_df <- dplyr::tibble(value_col = column_vector)
          
          category_summary_data <- temp_df %>%
            dplyr::filter(!is.na(value_col)) %>%
            dplyr::count(value_col, name = "n_cat") %>%
            dplyr::mutate(perc_cat = (n_cat / sum(n_cat)) * 100) %>%
            # Order categories by name for consistency
            dplyr::arrange(as.character(value_col)) 

          category_strings <- category_summary_data %>%
            dplyr::mutate(text = sprintf("%s: %d (%.1f%%)", as.character(value_col), n_cat, perc_cat)) %>%
            dplyr::pull(text)
          
          summary_str <- paste(category_strings, collapse = "; ")
        } else {
          summary_str <- "N/A (empty qualitative)"
        }
      } 
      # 2. Numeric columns
      else if (is.numeric(column_vector)) {
        if (non_na_count > 0) {
          if (col_name %in% mean_sd_vars) {
            mean_val <- mean(column_vector, na.rm = TRUE)
            sd_val <- sd(column_vector, na.rm = TRUE)
            if (is.na(sd_val)) { # sd can be NA if only one non-NA value or all identical
              summary_str <- sprintf("%.2f (SD N/A)", mean_val)
            } else {
              summary_str <- sprintf("%.2f \U00B1 %.2f", mean_val, sd_val) # Using Unicode for ±
            }
          } else if (col_name %in% median_iqr_vars) {
            median_val <- median(column_vector, na.rm = TRUE)
            q1_val <- quantile(column_vector, 0.25, na.rm = TRUE, names = FALSE)
            q3_val <- quantile(column_vector, 0.75, na.rm = TRUE, names = FALSE)
            summary_str <- sprintf("%.2f [%.2f - %.2f]", median_val, q1_val, q3_val)
          } else { # Default for numeric columns not specified in either list i.e median
            median_val <- median(column_vector, na.rm = TRUE)
            q1_val <- quantile(column_vector, 0.25, na.rm = TRUE, names = FALSE)
            q3_val <- quantile(column_vector, 0.75, na.rm = TRUE, names = FALSE)
            summary_str <- sprintf("%.2f [%.2f - %.2f]", median_val, q1_val, q3_val)
          }
        } else {
          summary_str <- "N/A (empty numeric)"
        }
      }
      # 3. Other types (should ideally not happen with typical data frames)
      else {
        summary_str <- "N/A (unsupported type)"
      }
      
      dplyr::tibble(
        Original_Column = col_name,
        Summary = summary_str,
        Non_NA_Count = as.integer(non_na_count)
      )
    })
  }

  # Write the summary tibble to a TSV file
  readr::write_tsv(summary_tibble, output_tsv_path)
  
  message(paste("Summary table successfully saved to", output_tsv_path))
  # Optionally return the tibble for use in R environment
  return(summary_tibble)
}
```

```{r}
hcmr_summary_output <- generate_summary_table_R(summary_tb, 'OUTPUT/HCMR/hcmr_summary_nontransformedCMR.tsv', 
                         mean_sd_vars=c(pp, 'bsa, cmrdiabp'),
                         median_iqr_vars = c('age','bmi',c('wallthkmax',hty,func,fibr)))
```

# Multivariate Regression Analysis

This performs multivariate linear regression (for continuous phenotypes) against the phenotype of interest + covariates of age/sex/BMI/BSA (if non-indexed).
You have to be careful about linear regression assumptions as well so some traits e.g LGE and the biomarkers need to be normalised.

The assumptions of multivariate linear regression include:

- Lack of multicollinearity (addressed by assessing VIF)
- Multivariate normality i.e residuals normally distributed (addressed by evaluating QQplots of residuals)
- Homoscedasticity of residuals (addressed by evaluating scale-location plot: flat horizontal line with equally spread points expected)
- Linearity (addressed by evaluating scatterplot of residuals vs. predicted values: flat horizontal line expected)

```{r multivariate_assessor_function}

multivariate_assessor <- function(input_tb, response_var, predictor_vars, covars=c('age','gender','bmi','cmrdiabp','pc1','pc2','pc3','pc4','pc5', 'sarcomere'),cont_or_discrete_response='cont', model_return=F, summary_print=T, ivnormalise=F){ #Assume already rank-based inverse normalised
  
  formula <- str_c(response_var, '~',
                   str_c(covars, collapse='+'),'+',
                   str_c(predictor_vars,collapse='+'))
  
  input_tb <- filter(input_tb, !is.na(!!sym(response_var))) #Exclude rows with NA values in the response variable
  
  if(cont_or_discrete_response=='cont'){
    model <- lm(formula, data=input_tb)
    
      if(isTRUE(ivnormalise)){
    input_tb[[response_var]] <- RNOmni::RankNorm(input_tb[[response_var]])
  }
    
  } else if (cont_or_discrete_response=='discrete'){
    model <- glm(formula,data=input_tb, family='binomial' )
  }
  
   if(isTRUE(summary_print)){
      print(str_c('Phenotype',response_var, 'tested for association with', str_c(predictor_vars, collapse=', '),sep=' '))
      print(summary(model))
   }
  
  #Evaluate the assumptions of multiple linear regression
  
  #Check for multicollinearity (using variance inflation factor)
  if(sum(vif(model)>5)>=1){ #If any of the coefficients have VIF greater than 5
    print('Variance inflation factor >5 detected in the model suggests multicollinearity')
    stop()
  }
  
  #Plot diagnostic plots including
  ## Residuals vs. predicted (to evaluate linearity assumption)
  ## Scale-location plot to evaluate homoscedasticity
  ## Residuals QQplot to evaluate normality of residuals
  ## Outliers/high-leverage points via studentised residuals vs. leverage plot 
  par(mfrow = c(2, 2))
  plot(model)
  
  pvals <- summary(model)$coefficients
  adjusted_rsq <- summary(model)$adj.r.squared
  
  output_tb <- as_tibble(pvals) %>%
    bind_cols('Predictor'=rownames(pvals)) %>%
    mutate('Pheno'=response_var) %>%
    select(Pheno,Predictor, everything()) %>%
    filter(Predictor %in% predictor_vars) 
  colnames(output_tb) <- c('Pheno','Predictor','Estimate','SE','tvalue','pval')
  
  output_tb <- output_tb %>% 
    mutate(upperSE = Estimate + SE,
           lowerSE = Estimate - SE) %>%
    select(-tvalue)
  
  if (isTRUE(model_return)){
    return(model)
  } else{
    return(output_tb)
  }
}
```

```{r summary_plotter}

summary_plotter <- function(input_tb,output_label, output_path='OUTPUT/HCMR/crosspheno_assoc/summary_plots/', pval='bonferroni',confint=F, se=F,test='t-Test', xlabel='Change in Phenotype per Unit Increase in Predictor'){
  
  if(length(unique(input_tb$Pheno))>8){
    Palette <- 'Paired'
  } else {
    Palette <- 'Dark2'
  }
  
  if (pval=='fdr'){
    input_tb <- input_tb %>%
    mutate(pval = p.adjust(pval, 'fdr'))
  } 
 
  summary_cont_plot <- ggplot(input_tb, aes(x=Estimate, y=-log10(pval), col=Label, shape=Label2))+
    geom_vline(xintercept=0, linetype='dashed')+
    geom_point()+
    scale_colour_brewer(palette=Palette)+
    labs(col='Phenotype',shape='Predictor')+
    scale_x_continuous(n.breaks=10)+
    xlab(xlabel)+
    ylab('-log10(p-value)')+
    labs(title=str_wrap(str_c('Summary plot of ', test,' tests for ',length(unique(input_tb$Pheno)), ' phenotypes across ', length(unique(input_tb$Predictor)), ' predictors')))
  
  if (pval=='bonferroni'){
    
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(0.05/nrow(input_tb)), linetype='dashed')+
      labs(caption=str_c('Bonferroni-corrected p-value threshold = ',signif(0.05/nrow(input_tb),3)))
  } else if (pval =='fdr'){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(0.05), linetype='dashed')+
      ylab('-log10(Adjusted p-value)')+
      labs(caption='FDR correction applied for multiple testing burden')
  }
  
  if(isTRUE(confint)){
    summary_cont_plot <- summary_cont_plot +
      geom_errorbar(aes(xmin=lowerCI, xmax=upperCI),alpha=0.5)
    
    append <- '_confint'
  } else if(isTRUE(se)){
      summary_cont_plot <- summary_cont_plot +
      geom_errorbar(aes(xmin=lowerSE, xmax=upperSE),alpha=0.5)
    
    append <- '_SE'
    }
    else{
    append <- ''
  }
  
  print(summary_cont_plot)
  ggsave(str_c(output_path,test,'_',output_label,'_summary_plot_',pval,append,'.png'),summary_cont_plot,dpi=600, width=12, height=6)
  
  return(input_tb)
}


```


## Group 1: Cardiac Fibrosis; Plasma Proteins; CMR Function/Hypertrophy Triangle

Note that when using groups of predictors, you should remove correlated predictors due to multicollinearity.

So there is some relationship likely between these three factors so assess:

1. Fibrosis ~ plasma proteins
2. Functional measures ~ plasma proteins
3. Hypertrophy measures ~ plasma proteins

### Joint CMR vs. Plasma Proteins Association Analysis

This evaluates in a joint model (i.e all together).

```{r g1_comp_vs_pp}
# #Fibrosis ~ plasma proteins
# print('Evaluating fibrosis ~ plasma')
# fibr_vs_pp <- map(fibr,~multivariate_assessor(transformed_tb,.x,pp))%>% bind_rows()
# 
# #Evaluate separately for LGE_pres due to discrete nature
# lge_pres_vs_pp <- multivariate_assessor(transformed_tb, 'lge_pres',pp,cont_or_discrete_response = 'discrete')
# fibr_vs_pp <- bind_rows(fibr_vs_pp, lge_pres_vs_pp)
# rm(lge_pres_vs_pp)
# 
# #Functional ~ plasma proteins
# print('Evaluating functional ~ plasma')
# func_vs_pp <- map(func,~multivariate_assessor(transformed_tb,.x,pp))%>% bind_rows()
# 
# #Hypertrophy ~ plasma proteins
# print('Evaluating hypertrophy ~ plasma')
# hty_vs_pp <- map(hty,~multivariate_assessor(transformed_tb,.x,pp))%>% bind_rows()
# 
# #Evaluate separately for wallthkmax due to need for BSA covariate
# wallthkmax_vs_pp <- multivariate_assessor(transformed_tb, 'wallthkmax',pp,covars=c('age','gender','bmi','bsa','pc1','pc2','pc3','pc4','pc5'))
# hty_vs_pp <- bind_rows(hty_vs_pp, wallthkmax_vs_pp)
# rm(wallthkmax_vs_pp)
# 
# all_vs_pp <- bind_rows(fibr_vs_pp, func_vs_pp, hty_vs_pp) %>%
#   left_join(select(label_tb,Variable,Label), by=c('Pheno'='Variable')) %>%
#   left_join(select(label_tb,Variable,Label2=Label), by=c('Predictor'='Variable'))
# rm(fibr_vs_pp, func_vs_pp, hty_vs_pp)
```

##### Summary Plot of Joint CMR Measures vs. Plasma Proteins

```{r g1_all_vs_pp_plot, fig.width=12, fig.height=6}

# summary_plotter(all_vs_pp,output_label='all_vs_pp_sarc',pval='fdr',se=F)
# summary_plotter(all_vs_pp,output_label='all_vs_pp_sarc',pval='fdr',se=T)

```

### Marginal CMR vs. Plasma Proteins Association Analysis

I also analyse each plasma protein individually via marginal models (hence CMR measure ~ covariates + plasma protein z).

```{r marginal_cmr_vs_pp}
#Fibrosis ~ plasma proteins
print('Evaluating fibrosis ~ plasma marginally')
fibr_vs_pp_combinations <- as.list(expand_grid(fibr,pp))
fibr_vs_pp_marginal <- map2(fibr_vs_pp_combinations[[1]],fibr_vs_pp_combinations[[2]],~multivariate_assessor(transformed_tb,.x,.y))%>% bind_rows()
rm(fibr_vs_pp_combinations)

#Evaluate separately for LGE_pres due to discrete nature
# lge_pres_vs_pp_combinations <- as.list(expand_grid('lge_pres',pp))
# lge_pres_vs_pp_marginal <- map2(lge_pres_vs_pp_combinations[[1]],lge_pres_vs_pp_combinations[[2]],~multivariate_assessor(transformed_tb,.x,.y,cont_or_discrete_response = 'discrete'))%>% bind_rows()
# fibr_vs_pp_marginal <- bind_rows(fibr_vs_pp_marginal, lge_pres_vs_pp_marginal)
# rm(lge_pres_vs_pp_marginal)

#Functional ~ plasma proteins
print('Evaluating functional ~ plasma marginally')
func_vs_pp_combinations <- as.list(expand_grid(func,pp))
func_vs_pp_marginal <- map2(func_vs_pp_combinations[[1]],func_vs_pp_combinations[[2]],~multivariate_assessor(transformed_tb,.x,.y))%>% bind_rows()

#Hypertrophy ~ plasma proteins
print('Evaluating hypertrophy ~ plasma marginally')
hty_vs_pp_combinations <- as.list(expand_grid(hty,pp))
hty_vs_pp_marginal <- map2(hty_vs_pp_combinations[[1]],hty_vs_pp_combinations[[2]],~multivariate_assessor(transformed_tb,.x,.y))%>% bind_rows()

#Evaluate separately for wallthkmax due to need for BSA covariate
wallthkmax_vs_pp_combinations <- as.list(expand_grid('wallthkmax',pp))
wallthkmax_vs_pp_marginal <- map2(wallthkmax_vs_pp_combinations[[1]],wallthkmax_vs_pp_combinations[[2]],~multivariate_assessor(transformed_tb,.x,.y,covars=c('age','gender','bmi','cmrdiabp','pc1','pc2','pc3','pc4','pc5', 'sarcomere', 'bsa')))%>% bind_rows()
hty_vs_pp_marginal <- bind_rows(hty_vs_pp_marginal, wallthkmax_vs_pp_marginal)
rm(wallthkmax_vs_pp_marginal)

all_vs_pp_marginal <- bind_rows(fibr_vs_pp_marginal, func_vs_pp_marginal, hty_vs_pp_marginal) %>%
  left_join(select(label_tb,Variable,Label), by=c('Pheno'='Variable')) %>%
  left_join(select(label_tb,Variable,Label2=Label), by=c('Predictor'='Variable'))
rm(fibr_vs_pp_marginal, func_vs_pp_marginal, hty_vs_pp_marginal)

```

Add colour and shape variables for plotting to the tb.

```{r}
#Remove lge_presence
all_vs_pp_marginal <- all_vs_pp_marginal %>%
  filter(Pheno != 'lge_pres')

all_vs_pp_marginal <- all_vs_pp_marginal %>%
  mutate(phenogroup = case_when(Pheno %in% func ~ 'Contractile Function',
                              Pheno %in% fibr ~ 'Fibrosis',
                              Pheno %in% hty | Pheno == 'wallthkmax' ~'Hypertrophy',
                              T~'other'))
shape_mapping <- c('lvmassi'=17,'wallthkmax'=18,  # Hypertrophy shapes
                    'lge_total' = 15, 'ecvfwhole' = 16,   # Fibrosis shapes
                    'lvef' = 0, 'lvs_ellmean' =1, 'lvs_eccmean'=2, 'lvs_errmean' = 3, 'ci' = 4)   # Contractility shapes

all_vs_pp_marginal <- all_vs_pp_marginal %>%
  mutate(p_shape = shape_mapping[Pheno])

write_tsv(all_vs_pp_marginal,'OUTPUT/HCMR/crosspheno_assoc/all_vs_pp_marginal.tsv')

```

##### Summary Plot of Marginal CMR Measures vs. Plasma Proteins

```{r summary_marginal_cmr_vs_pp}
summary_plotter(all_vs_pp_marginal,output_label='all_vs_pp_marginal_sarc',pval='fdr',se=F,xlabel='Change in Phenotype per SD of Plasma Protein')
summary_plotter(all_vs_pp_marginal,output_label='all_vs_pp_marginal_sarc',pval='fdr',se=T,xlabel='Change in Phenotype per SD of Plasma Protein')
```

You can also plot out each individual plot on a per-biomarker basis, keeping the multiple testing correction the same by using the adjusted p-values over their entirety.

```{r}
summary_plotter_individualpredictor <- function(input_tb,predictor_of_interest, output_path='OUTPUT/HCMR/crosspheno_assoc/summary_plots/', pval='bonferroni',confint=F, se=F,test='t-Test', xlabel='Change in Phenotype per Unit Increase in Predictor', ylimits='', xlimits=''){

  # if(length(unique(input_tb$Pheno))>8){
  #   Palette <- 'Paired'
  # } else {
  #   Palette <- 'Dark2'
  # }

  if (pval=='fdr'){
    input_tb <- input_tb %>%
    mutate(pval = p.adjust(pval, 'fdr'))
    p_threshold <- 0.05
  } else if (pval=='bonferroni'){
    p_threshold <- 0.05/nrow(input_tb) #Define the multiple testing corrected p-value threshold for ALL the phenotypes
  }

  input_tb <- input_tb %>%
    filter(Predictor == predictor_of_interest)
  
  input_tb$p_shape <- setNames(input_tb$p_shape, input_tb$Label)

  summary_cont_plot <- ggplot(input_tb, aes(x=Estimate, y=-log10(pval))) +
    geom_vline(xintercept=0, linetype='dashed') +
    geom_point(aes(col=phenogroup, shape=Label), size=4) +
    scale_colour_brewer(palette='Set2')+
    scale_shape_manual(values = input_tb$p_shape) +  # Fix: Assign shape values to Labels
    labs(col='Phenotype Group', shape='Phenotype') +
    scale_x_continuous(n.breaks=10) +
    xlab(xlabel) +
    ylab('-log10(p-value)') +
    labs(title=str_wrap(str_c('Summary plot of ', test, ' tests for ',
                              length(unique(input_tb$Pheno)), ' phenotypes across ',
                              length(unique(input_tb$Predictor)), ' predictors')))
  # +
  #   theme(legend.position = "bottom",
  #         legend.box = "vertical",
  #         legend.box.just = "left" )+ 
  #   guides(col=guide_legend(ncol=1, order=1), 
  #          shape = guide_legend(ncol = 2, order=2))

  if (pval=='bonferroni'){

    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(p_threshold), linetype='dashed')+
      labs(caption=str_c('Bonferroni-corrected p-value threshold = ',signif(p_threshold,3)))
  } else if (pval =='fdr'){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(p_threshold), linetype='dashed')+
      ylab('-log10(Adjusted p-value)')+
      labs(caption='FDR correction applied for multiple testing burden')
  }

  if(isTRUE(confint)){
    summary_cont_plot <- summary_cont_plot +
      geom_errorbar(aes(xmin=lowerCI, xmax=upperCI),alpha=0.5)

    append <- '_confint'
  } else if(isTRUE(se)){
      summary_cont_plot <- summary_cont_plot +
      geom_errorbar(aes(xmin=lowerSE, xmax=upperSE),alpha=0.5)

    append <- '_SE'
    }
    else{
    append <- ''
    }
  
  if(!identical(ylimits,'')){
    summary_cont_plot <- summary_cont_plot + 
        scale_y_continuous(limits=ylimits)
  }
  
  if(!identical(xlimits,'')){
    summary_cont_plot <- summary_cont_plot + 
        scale_x_continuous(n.breaks=10, limits=xlimits)
  }
  
  append <- str_c(append, str_c(xlimits, collapse=''), '_', str_c(ylimits, collapse=''))

  print(summary_cont_plot)
  ggsave(str_c(output_path,test,'_',predictor_of_interest,'_indivpredictor_plot_',pval,append,'.png'),summary_cont_plot,dpi=600, width=12, height=6)
}
```

Now run over all the pp.

```{r}
walk(pp, ~summary_plotter_individualpredictor(all_vs_pp_marginal, ., pval='fdr',se=T,xlabel='Change in Phenotype per SD of Plasma Protein'))

walk(pp, ~summary_plotter_individualpredictor(all_vs_pp_marginal, ., pval='fdr',se=T,xlabel='Change in Phenotype per SD of Plasma Protein',ylimits=c(0,150), xlimits =c(-0.5,0.6)))

walk(pp[!pp %in% c('NTproBNP','TnTStat')], ~summary_plotter_individualpredictor(all_vs_pp_marginal, ., pval='fdr',se=T,xlabel='Change in Phenotype per SD of Plasma Protein',ylimits=c(0,4), xlimits =c(-0.12,0.1)))
```

# Addendum - GWAS Prep

This prepares the .pheno and .covar files for GWAS of HCMR plasma proteins using REGENIE.

```{r}
#Import in the total_df_notransform.rds containing phenotype information
total_df_notransform <- readRDS('../../EDA_HCMR/output/data/total_notransform_df.rds')

#Import in the 10PCs generated in HCMR only
eigenvectors <- read_tsv('popgen/2_gwas/data/hcmr/HCMR_HARDCALLS/eigenvectors_hcmr_10pcs') %>%
  rename_with(~str_replace(.,'U','pc'),contains('U'))

#Join to total_df_notransform and plot a quick visualisation of PC1 vs. 2 mapped onto self-reported ancestry

total_df_notransform <- total_df_notransform %>%
  left_join(select(eigenvectors,-FID), by=c('HCR_IDs'='IID'))

total_df_notransform <- filter(total_df_notransform,!is.na(pc1)) #Filters out HCR00011 who fails the hard-called genotypes QC filter

pcplot <- ggplot(total_df_notransform, aes(x=pc1,y=pc2))+
  geom_point(aes(col=race))
print(pcplot) #Looks like good separation of self-reported ancestry

```

```{r}
regenie_pheno_covar_prep <- function(input_tb, phenotypes_of_interest, output_filepath, samplefile_path, covariates=c('sex','age','height','weight',str_c('pc',seq(1,10))) ){

  input_tb <- input_tb %>%
        dplyr::rename(ID=HCR_IDs, sex=gender)
  
  snptest_sample <- input_tb %>%
    select(sample_ID=ID, all_of(covariates), all_of(phenotypes_of_interest)) %>%
     mutate(sex=case_when(sex=='Male'~1,
                          sex=='Female'~0,
                          T~NA)) %>%
    mutate(across(.cols=everything(), as.character))
  
  sample_file <- read.table(samplefile_path, header=T) %>% #This is the original filepath for the .sample file corresponding to the .bgen file 
    select(sample_ID=1) %>%
    filter(sample_ID != 0) %>% #Filter out the first row which is a metadata row
    mutate(sample_ID = as.character(sample_ID))
  
  snptest_sample <- left_join(sample_file, snptest_sample, by='sample_ID') %>% #This is needed to keep the correct order of rows in the sample ID as the .bgen file for UKB imputed genotypes
    mutate(include=as.character(ifelse(sample_ID %in% input_tb$ID, T,F)))
  
  #snptest_sample <- filter(snptest_sample, include==T)
  
  regenie_covar <- snptest_sample %>%
    select(FID = sample_ID, IID=sample_ID, all_of(covariates))
  
  regenie_pheno<- snptest_sample %>%
    select(FID=sample_ID, IID=sample_ID, all_of(phenotypes_of_interest))
  
  #Also write sample inclusion file
  regenie_sample_inclusion <- snptest_sample %>%
    filter(include==T) %>%
    select(FID=sample_ID, IID=sample_ID)
  
    #Also write non-NA phenotype and non-NA covariate .incl file for FINEMAP/LDSTORE2
  nonNAphenocovar <- snptest_sample %>%
    filter(if_any(phenotypes_of_interest, ~!is.na(.))) %>%
    filter(if_all(covariates, ~!is.na(.))) %>%
    select(sample_ID)
  
  write_tsv(regenie_covar, str_c(output_filepath,'regenie_hcmr_covar.tsv'))
  write_tsv(regenie_pheno, str_c(output_filepath,'regenie_hcmr_pheno.tsv'))
  write_tsv(regenie_sample_inclusion, str_c(output_filepath, 'regenie_hcmr_sampleinclusion.tsv'), col_names = F)
  write_tsv(nonNAphenocovar, str_c(output_filepath, 'regenie_hcmr_sampleinclusion_nonNAphenocovars.incl'), col_names = F)
  
  return(list(regenie_pheno, regenie_covar))

}

hcmr_pp_regenie<- regenie_pheno_covar_prep(total_df_notransform,
                                                       c('NTproBNP','TnTStat','mmp1','timp1','cicp','gal3','st2'),
                                                       'popgen/2_gwas/data/hcmr/REGENIE/',
                                                       'popgen/2_gwas/data/hcmr/hcmr_htscores_v2_essentialcols.sample')
```

## Misc.

### NTproBNP Distribution comparison between HCMR and UKB

```{r}
#Assume you have run the import code in 2_UKB_CrossPheno_AssocAnalysis.Rmd

merged_ntprobnp <- bind_rows(select(total_df_notransform, NTproBNP) %>% mutate(group='HCMR') %>% filter(!is.na(NTproBNP)) %>% mutate(NTproBNP = RNOmni::RankNorm(NTproBNP)), select(ukb_20240209_select_pp, NTproBNP) %>% mutate(group='UKB'))

dist <- ggplot(merged_ntprobnp,aes(x=NTproBNP))+
  geom_density(aes(fill=group, col=group),alpha=0.5)+
  scale_colour_brewer(palette='Set2')+
  scale_fill_brewer(palette='Set2')

print(dist)
```

### Endophenotype Comparison within Sarc+ve via DBP

This evaluates sarcomere +ve individuals to assess correlations between DBP and endophenotypes after adjusting for covariates.

```{r}
df <- transformed_tb %>%
  filter(sarcomere==T)

func_vs_dbp_marginal <- map(func,~multivariate_assessor(filtered_tb,.x,'cmrdiabp', covars=c('age','gender','bmi','pc1','pc2','pc3','pc4','pc5')))%>% bind_rows()
hty_vs_dbp_marginal <- map(hty,~multivariate_assessor(filtered_tb,.x,'cmrdiabp', covars=c('age','gender','bmi','pc1','pc2','pc3','pc4','pc5')))%>% bind_rows()
fibr_vs_dbp_marginal <- map(fibr,~multivariate_assessor(filtered_tb,.x,'cmrdiabp', covars=c('age','gender','bmi','pc1','pc2','pc3','pc4','pc5')))%>% bind_rows()
maxlvwt_vs_dbp_marginal <- multivariate_assessor(df, 'wallthkmax','cmrdiabp', covars=c('age','gender','bmi','pc1','pc2','pc3','pc4','pc5','bsa'))

all_vs_dbp_marginal <- bind_rows(fibr_vs_dbp_marginal, func_vs_dbp_marginal, hty_vs_dbp_marginal, maxlvwt_vs_dbp_marginal) %>%
  left_join(select(label_tb,Variable,Label), by=c('Pheno'='Variable')) %>%
  left_join(select(label_tb,Variable,Label2=Label), by=c('Predictor'='Variable'))
rm(fibr_vs_dbp_marginal, func_vs_dbp_marginal, hty_vs_dbp_marginal, maxlvwt_vs_dbp_marginal)

summary_plotter(all_vs_dbp_marginal,output_label='all_vs_dbp_marginal_sarcposonly',pval='fdr',se=F,xlabel='Change in Phenotype per 1mmHg of DBP')
summary_plotter(all_vs_dbp_marginal,output_label='all_vs_dbp_marginal_sarcposonly',pval='fdr',se=T,xlabel='Change in Phenotype per 1mmHg of DBP')
```

### NTproBNP and Troponin Check in Normal LVEF vs. Decreased LVEF Patients

```{r}

#Convert LVEF and NTproBNP and TnTStat back to non-normalised form
lvef_stratifier_tb <- left_join(select(transformed_tb,-lvef), select(total_df_notransform_rare, HCR_IDs, lvef), by='HCR_IDs')

lvef_stratifier_check <- function(input_tb, pheno_of_interest,lvef_cutoff=50, plot_output_path = '../../EDA_HCMR/output/2_CrossPheno_AssocAnalysis/hcmr/misc/', suffix=''){
  
  input_tb <- input_tb %>%
    filter(!is.na(lvef)) %>%
    filter(!is.na(eval(parse(text=pheno_of_interest)))) %>%
    mutate(group = ifelse(lvef < lvef_cutoff, str_c('< ',lvef_cutoff, '% LVEF'),str_c('>= ',lvef_cutoff, '% LVEF') ))
  
  wilcox_test <- wilcox.test(filter(input_tb, group==str_c('< ',lvef_cutoff, '% LVEF'))[[pheno_of_interest]],
                   filter(input_tb, group==str_c('>= ',lvef_cutoff, '% LVEF'))[[pheno_of_interest]],
                   alternative='two.sided', conf.int=T) #Welch t-test with unequal variance
  
  density_comp <- ggplot(input_tb, aes(x=eval(parse(text=pheno_of_interest)), fill=group))+
    geom_density(alpha=0.5)+
    ylab('Density')+
    xlab(pheno_of_interest)+
    labs(title=str_wrap(str_c('Density comparison with stratification by LVEF at ',lvef_cutoff, '% for phenotype ',pheno_of_interest)), subtitle = str_c('Non-NA individuals = ',nrow(input_tb)),
         fill='Group',
         caption= str_c('Wilcoxon p-value = ',signif(wilcox_test$p.value,3), ' with 95%CI = ', signif(wilcox_test$conf.int[[1]],3), ' - ', signif(wilcox_test$conf.int[[2]],3), ' SD'))
  
  print(density_comp)
  ggsave(str_c(plot_output_path, 'lvef_stratification_density_thresh',lvef_cutoff,'_',pheno_of_interest,suffix,'.png'),dpi=600)
  
  return(input_tb)
}

pp_of_interest <- c('NTproBNP','TnTStat')

walk(pp_of_interest, ~lvef_stratifier_check(lvef_stratifier_tb, .)) #50% LVEF threshold
walk(pp_of_interest, ~lvef_stratifier_check(lvef_stratifier_tb, ., lvef_cutoff=60)) #60% LVEF threshold

```

This identifies the median NTproBNP and TnTStat levels in the HCMR individuals with LVEF >50% and <50%.

```{r}
#Check the non-LVEF NTproBNP levels on raw scale
#Convert LVEF and NTproBNP and TnTStat back to non-normalised form
nontransformed_lvef_tb <- left_join(select(transformed_tb,-lvef,-NTproBNP, -TnTStat), select(total_df_notransform_rare, HCR_IDs, lvef, NTproBNP, TnTStat), by='HCR_IDs')

lvef50_cutoff_pp <- map(pp_of_interest, ~lvef_stratifier_check(nontransformed_lvef_tb, ., suffix = '_nontransformedpp')) #50% LVEF threshold
map(pp_of_interest, ~lvef_stratifier_check(nontransformed_lvef_tb, ., lvef_cutoff=60, suffix = '_nontransformedpp')) #60% LVEF threshold

#Find the median and upper/lower quartiles of HCMR individuals with LVEF >50%
print(str_c('Median NTproBNP = ',
  median(filter(lvef50_cutoff_pp[[1]], lvef >=50)$NTproBNP),' [', 
  quantile(filter(lvef50_cutoff_pp[[1]], lvef >=50)$NTproBNP,0.25),' - ',
  quantile(filter(lvef50_cutoff_pp[[1]], lvef >=50)$NTproBNP,0.75),'] ',
  'pg/mL'))

print(str_c('Median TnT = ',
  median(filter(lvef50_cutoff_pp[[1]], lvef >=50)$TnTStat),' [', 
  quantile(filter(lvef50_cutoff_pp[[1]], lvef >=50)$TnTStat,0.25),' - ',
  quantile(filter(lvef50_cutoff_pp[[1]], lvef >=50)$TnTStat,0.75),'] ',
  'ng/L'))

#Find the median and upper/lower quartiles of HCMR individuals with LVEF >50%
print(str_c('Median NTproBNP = ',
  median(filter(lvef50_cutoff_pp[[1]], lvef <50)$NTproBNP),' [', 
  quantile(filter(lvef50_cutoff_pp[[1]], lvef <50)$NTproBNP,0.25),' - ',
  quantile(filter(lvef50_cutoff_pp[[1]], lvef <50)$NTproBNP,0.75),'] ',
  'pg/mL'))

print(str_c('Median TnT = ',
  median(filter(lvef50_cutoff_pp[[1]], lvef <50)$TnTStat),' [', 
  quantile(filter(lvef50_cutoff_pp[[1]], lvef <50)$TnTStat,0.25),' - ',
  quantile(filter(lvef50_cutoff_pp[[1]], lvef <50)$TnTStat,0.75),'] ',
  'ng/L'))


```



