---
title: "UKB_PlasmaProteomics_TimeInstancing_SensitivityAnalysis"
author: "Jonathan Chan"
date: "2025-04-01"
output: html_document
---

# Time Instancing Sensitivity Analysis

This performs sensitivity analysis on the time instancing i.e the max. lag time from blood sample collection before which diagnosis of HCM is assumed to occur.

```{r}
library(tidyverse)
library(tictoc)
library(ggrepel)
library(ggsignif)
library(car)
library(readxl)
library(limma)
library(ggnewscale)
library(survival)

theme_set(theme_classic(base_size=14))

date <- as.POSIXlt(Sys.time(), 'UTC')
date <- strftime(date, "%Y%m%dT%H:%M:%S%z")
date <- str_match(date,'^(\\d+)T')[,2]
```

## Time-Instancing Case Calculator

This is a function used to compute the cases based on the lag time for HCM only.
We assume 0 time lag for individuals who are HF or T2D.

```{r}
hcm_case_instance_calculator <- function(input_tb, output_name, threshold_date='bcollection_date_i0', 
                                         lagtime=5, #Lag time defined in years
                                         output_folder='DATA/UKB/PROCESSED/5_TimeInstancing/') { 
  
  output_tb <- input_tb %>%
    mutate(case = ifelse(time_length(min_diag_date - eval(parse(text=threshold_date)), unit='year') <= lagtime |  #Assign 5Y status if their minimum date of diagnosis was within n years after attendance. NB This includes i0 cases because they have a time_length of negative because min_diag_Date < date_attend_i0
                                    time_length(death_date - eval(parse(text=threshold_date)), unit='year') <=lagtime, T,F)) #OR if they had mention of death with ICD10 code of interest & died within n years of attendance
    
  print(str_c('There are',nrow(filter(output_tb, case==T)), output_name, 'cases in the UKB who satisfy', lagtime, 'Y criteria', sep=' ' ))
  
  if(!dir.exists(output_folder)){dir.create(output_folder, recursive=T)}
  write_tsv(filter(output_tb, case==T)%>% select(ID), str_c(output_folder, 'ukb_rap_',output_name, '_20240209_',lagtime,'Ycases.tsv'))
  
  output_tb <- filter(output_tb, case==T)%>% select(ID) #Only output the filtered HCM cases
  
  return(output_tb)
}
```

## TB Builder

```{r}
overall_tb_builder <- function( ukb_cases, #HCM cases in UKB from the particular instance
                                hcm_instance, #Name of the instance
                                ukb_pp_list, #List of plasma proteins to compare
                              ukb_basic_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_age_sex_htwt_10pcs_dbp_t2d_smoking_cov.tsv',  
                                ukb_rcmm_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_rcmm_cov.tsv',
                                ukb_pp_filepath='DATA/UKB/RAP/PP/ukb_allpp_i0.tsv', 
                                include_nonCM_panels=T, rcmm_cov=T){
  
  #Import in the plasma protein data
  ukb_pp_data <- read_tsv(ukb_pp_filepath, col_types=rep('n',3006))%>% #All numeric columns
  select(-`...1`) %>%
  select(contains('instance_0')) %>% #It contains some instance 2 columns for some reason
  rename_with(~str_match(.,'\\.(.+)')[,2]) %>%
  filter(if_any(!'eid', ~!is.na(.))) %>%
  mutate(non_na_count = rowSums(!is.na(.))-1) %>% #-1 for eid is non-NA
  #filter(non_na_count > 2) %>%
  mutate(instance=0) %>%
  select(eid, instance, ukb_pp_list$ukb_fieldname_equivalent) #Filter for only plasma proteins in the ukb_pp_list
  
  #Import in covariate data
  ukb_20240209_snptest_covariates <- read_tsv(ukb_basic_cov_filepath)
  if(isTRUE(rcmm_cov)){ukb_20240209_rcmm_cov <- read_tsv(ukb_rcmm_cov_filepath)}
  
  #Combine the covariate + pp data
  ukb_pp_hcm_cov2 <- ukb_pp_data %>%
  mutate(hcm = ifelse(eid %in% ukb_cases$ID, T, F)) %>%
  left_join(ukb_20240209_snptest_covariates,by=c('eid'='ID')) %>%
  select(eid, instance, hcm:t2d, everything()) %>%
  relocate(bmi, .after='weight') %>% 
  filter(if_all(age:t2d, ~!is.na(.))) %>%  #Filter out individuals who have NA value for covariates 
  filter(if_any(all_of(ukb_pp_list$ukb_fieldname_equivalent), ~!is.na(.))) %>% #Filter out individuals who have no non-NA values for all panel proteins 
  select(where(~any(!is.na(.)))) #Filter for plasma proteins which actually have non-NA values 
  
  if(isTRUE(rcmm_cov)){
    ukb_pp_hcm_cov2 <- ukb_pp_hcm_cov2 %>%
    left_join(ukb_20240209_rcmm_cov,by=c('eid'='ID')) %>%
    select(eid, instance, hcm:t2d, Townsend:ldl, everything()) %>% #52092 individuals
    select(-hdl) %>%  #Ignore HDL for now due to loss of non-NA individuals
    filter(if_all(c(Townsend,eGFR,ldl), ~!is.na(.)))#Filter out individuals who have NA value for covariates
  }
  
  print(str_c('Number of HCM cases in instance ', hcm_instance, ' = ',sum(ukb_pp_hcm_cov2$hcm)))
  print(str_c('Number of total individuals with non-NA covariates + at least one non-NA plasma proteins in ', hcm_instance, ' = ',nrow(ukb_pp_hcm_cov2)))
  
  return(ukb_pp_hcm_cov2) #This technically doesn't have ranknorm normalisation
    
}
```

# Limma
```{r}
limma_dpe <- function(input_tb, protein, predictor_vars='hcm', covar=c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking'), ivnormalise=T){
  
  input_tb <- filter(input_tb, !is.na(!!sym(protein))) #Exclude rows with NA values in the response variable
  
  if(isTRUE(ivnormalise)){
    input_tb[[protein]] <- RNOmni::RankNorm(input_tb[[protein]])
  }
  
  #Separate into pheno_df for covariates and predictors and npx_df containing the NPX value for the protein over all individuals
  pheno_df <- select(input_tb, all_of(predictor_vars),all_of(covar)) %>%
    mutate(across(all_of(predictor_vars), ~factor(.)))
  npx_df <- select(input_tb, !!sym(protein))
  
  #Need to define the formula i.e model and the model matrix
  formula <- str_c('~0 +',
                   str_c(predictor_vars,collapse='+'),
                   '+',str_c(covar, collapse='+'))
  design <- model.matrix(eval(parse(text=formula)), pheno_df)
  
  #Need to create the matrix array for limma i.e rows = proteins; columns = samples via t() for transpose
  #Perform the model fitting using weighted least-squares
  lin_model <- lmFit(t(npx_df), design)
  
  #To obtain log-fold changes between groups of the predictor_var i.e hcmTRUE vs. hcmFALSE, you need to obtain the contrast of this fitted linear model
  if(predictor_vars=='hcm'){
      contr <- makeContrasts(hcmTRUE - hcmFALSE, levels = design)
  } else if (predictor_vars=='hf'){
    contr <- makeContrasts(hfTRUE - hfFALSE, levels = design)
  }else if (predictor_vars=='rare'){
    contr <- makeContrasts(rareTRUE - rareFALSE, levels = design)
  }
  
  # Estimate contrast for the protein of interest
  tmp <- contrasts.fit(lin_model, contr)
  
  # Empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other proteins towards the average standard error)
  tmp <- eBayes(tmp)
  
  #Return results for that particular protein
  top.table <- topTable(tmp, sort.by = "P", n = Inf)
  results <-  as_tibble(top.table) %>%
    select(-5) %>%
    mutate(pp=protein) %>%
    select(pp, everything())
  
  return(results)
}
```

## Summary Plotting
```{r}
summary_plotter_limma <- function(input_tb, output_label='ukb_allpanel', predictor = 'HCM status', output_path=str_c('PLOTS/UKB/1_Case_vs_Control/2_Limma/',instance), pval='bonferroni', fdr_threshold = 0.05, test='Empirical Bayes moderated t-test',xlabel='Log2(Fold-Change)', xlimits=c(-1,2),ylimits=c(0,40), specific_label=F){
  
  if (pval=='fdr'){
    input_tb <- input_tb %>%
      mutate(P.Value = p.adjust(P.Value, 'fdr'))
    
    pval_threshold <- fdr_threshold
  } else if(pval=='bonferroni'){
    pval_threshold <- 0.05/nrow(input_tb)
    
  } else if (is.numeric(pval)){ #i.e if defining own pvalue threshold to use
    pval_threshold <- pval
  }
  
  if(isFALSE(specific_label)){
    input_tb <- input_tb %>% #Only label the ones which reach significance
      mutate(Label=ifelse(P.Value < pval_threshold,Gene, ''),
           MTC_significant = ifelse(P.Value < pval_threshold, T, F)) %>%
      arrange(Label)
  } else {
    input_tb <- input_tb %>%
      mutate(MTC_significant = ifelse(Label=='', F, T)) %>%#Just for downstream code purposes to label some specific points
      arrange(Label)
  }

 
  #Base plot
  summary_cont_plot <- ggplot(input_tb, aes(x=logFC, y=-log10(P.Value), col=MTC_significant))+
    geom_vline(xintercept=0, linetype='dashed')+
    geom_point()+
    scale_colour_manual(name=ifelse(isFALSE(specific_label),'MTC Significant','PP of Interest'),values=c('FALSE'='grey','TRUE'='red'))+
    geom_text_repel(aes(label=Label), nudge_y=3 ,
                    #nudge_x=0.5,
                    force =5, max.overlaps=100)+
    labs(col='Phenotype',shape='Predictor')+
    scale_x_continuous(n.breaks=10, limits=xlimits)+
    scale_y_continuous(limits=ylimits)+
    xlab(xlabel)+
    ylab('-log10(p-value)')+
    labs(title=str_c('Summary plot of ', test,' tests for ',length(unique(input_tb$pp)), ' plasma proteins for ', predictor, ' predictor variable'))
  
  
  #Adjust the plot depending on the multiple testing correction
  if (pval=='bonferroni'){
    
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(0.05/nrow(input_tb)), linetype='dashed')+
      labs(caption=str_c('Bonferroni-corrected p-value threshold = ',signif(0.05/nrow(input_tb),3)))
    
  } else if (pval =='fdr'){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(pval_threshold), linetype='dashed')+
      ylab('-log10(Adjusted p-value)')+
      labs(caption=str_c( fdr_threshold*100, '% FDR correction applied for multiple testing burden'))
  } else if (is.numeric(pval)){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(pval_threshold), linetype='dashed')+
      labs(caption=str_c('Applied p-value threshold = ',signif(pval_threshold,3)))
  }
  
  print(summary_cont_plot)
  ggsave(str_c(output_path,output_label,'_summary_plot_',pval,'.png'),summary_cont_plot,dpi=600, width=12, height=6)
  
  #Also output a MAplot 
  ma_plot <- ggplot(input_tb, aes(x = log10(abs(AveExpr)), y = logFC)) +
    geom_point(aes(col= MTC_significant), alpha=0.5) +
    geom_text_repel(aes(label=Label))+
    labs(x = "Log10(Mean Normalised NPX)", y = xlabel) +
    scale_colour_manual(name='MTC Significant',values=c('FALSE'='grey','TRUE'='red'))+
    labs(title=str_c('MA plot for ', test, ' in UKB plasma proteins CM panel'))

  print(ma_plot)
  ggsave(str_c(output_path,output_label,'_ma_plot_',pval,'.png'),ma_plot,dpi=600, width=9, height=6)
  
  return(input_tb)

}
```


## Main
```{r}
#Earliest collection time for HCM case in UKB = ~ 2006 and end of dataset = 2024 so 18 years is max
lagtimes <- seq(0,17)
ukb_pp_list <- read_tsv('DATA/UKB/PROCESSED/ukb_pp_list_allpanel.tsv')
hf_eid <- read_tsv('DATA/UKB/PROCESSED/1_I0/ukb_rap_HF_20240209_i0cases.tsv')

hcm_diag_tb <- readRDS('DATA/UKB/PROCESSED/3_TOTAL/ukb_diagnosis_dates_minimum.rds')
bcollection_date <- read_tsv('DATA/UKB/RAP/Cases/2024_02_09_allUKB_BloodCollection_Time.tsv')%>%
  dplyr::rename(ID=eid, bcollection_date_i0=p21842_i0) %>%
  mutate(bcollection_date_i0=as_date(bcollection_date_i0))
hcm_diag_tb <- left_join(hcm_diag_tb, bcollection_date, 'ID')

main <- function(lagtime,  #Lagtime to test in years
                  ukb_pp_list, #PPs to include in analysis,
                  hcm_diag_tb, #This is the 
                  hf_eid, #IDs of HF cases (i0)
                  ukb_basic_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_age_sex_htwt_10pcs_dbp_t2d_smoking_cov.tsv',  
                  ukb_rcmm_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_rcmm_cov.tsv',
                  ukb_pp_filepath='DATA/UKB/RAP/PP/ukb_allpp_i0.tsv', 
                  pp_scaling='ranknorm',
                  rcmm_cov=T,
                  include_nonCM_panels=T,
                  cc_output_basepath = 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/',
                  limma_run=T){
  
  hcm_cases <- hcm_case_instance_calculator(hcm_diag_tb, 'HCM', lagtime=lagtime)
  
  ukb_pp_cov_hcm <- overall_tb_builder(hcm_cases, #HCM cases in UKB from the particular instance
                                str_c(lagtime, 'Y', collapse = ''), #Name of the instance
                                ukb_pp_list, #List of plasma proteins to compare
                                ukb_basic_cov_filepath,  
                                ukb_rcmm_cov_filepath,
                                ukb_pp_filepath, 
                                include_nonCM_panels, rcmm_cov)
  
  pp <- ukb_pp_list$ukb_fieldname_equivalent[ukb_pp_list$ukb_fieldname_equivalent %in% colnames(ukb_pp_cov_hcm)] %>%  #In case any PP are removed due to lack of non-NA values
  unique()
  
  #Run case-control analysis for HCM (including HF) vs. non-HF (i0) and non-HCM (lagtime) controls
  #Only run this CC analysis (and not subgroup ones) because limma is a bit slow
  
  cc_tb <- ukb_pp_cov_hcm %>%
    mutate(hcm = case_when(eid %in% hcm_cases$ID ~ T, #HCM cases independent of HF status
                           !eid %in% hcm_cases$ID & !eid %in% hf_eid$ID ~ F, #Non-HF non-HCM controls
                           T~NA)) %>% #non-HCM and HF 
    filter(!is.na(hcm))
  
  print(str_c('At lagtime ', lagtime, 'Y, number of HCM cases (incl. those with HF) = ', sum(cc_tb$hcm==T), ' and non-HCM/HF controls = ', sum(cc_tb$hcm==F)))
  

  if (isTRUE(rcmm_cov)){  #If using the extended covariate list suggested by Rob Clarke and Mohsen
    extra_folders <- 'rcmm_cov/'
    covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking','Townsend','eGFR', 'ldl')
  } else {
    covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking')
  }
  if(isTRUE(include_nonCM_panels)){ extra_folders <- str_c(extra_folders, 'allpanels/')} 
  
  path <- str_c(cc_output_basepath,'/HCM/', extra_folders) #Define the folder for output of plots and tabular data
  if(!dir.exists(path)){dir.create(path, recursive=T)}
  
  if(isTRUE(limma_run)){
    limma_hcm <- map(pp, ~limma_dpe(cc_tb,., covar=covars)) %>% bind_rows()
    
    limma_hcm <- left_join(limma_hcm, select(ukb_pp_list, Gene=Target, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent')) %>% unique()
    write_tsv(limma_hcm, str_c(path,lagtime, 'Y_lag_limma_results.tsv'))
    print(str_c('Written output limma_results.tsv for lagtime ', lagtime, 'Y'))
    
    pp_vs_hcm_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm, output_label = str_c(lagtime, 'Y_lag_allpp_vs_hcm'),predictor = 'HCM status',output_path=path, xlimits=c(-1,2),ylimits=c(0,50)) 
    
    #Also for 5%FDR and 1%FDR
    pp_vs_hcm_marginal_fdr5_summarytb_limma <- summary_plotter_limma(limma_hcm, output_label = str_c(lagtime, 'Y_lag_allpp_vs_hcm'),predictor = 'HCM status',output_path=path, pval = 'fdr', xlimits=c(-1,2),ylimits=c(0,50)) 
    
    rm(ukb_pp_cov_hcm, path)
  
  return(limma_hcm)
  } else{
    return(c(sum(cc_tb$hcm==T), sum(cc_tb$hcm==F)))
  }
  

}

# walk(lagtimes, ~main(., ukb_pp_list, hcm_diag_tb, hf_eid)) #Run limma
perlagtime_tb <- map(lagtimes, ~main(., ukb_pp_list, hcm_diag_tb, hf_eid, limma_run=F)) #Don't run limma and only output the tbs
```

# Case/Control Number for each Lagtime

```{r}
# case_numbers <- map_int(perlagtime_tb, ~lambda x(sum(x$hcm==T))) %>% bind_cols('Lagtime' = seq(0,17))
# control_numbers <- map_int(perlagtime_tb, ~lambda x(sum(x$hcm==F))) %>% bind_cols('Lagtime' = seq(0,17))

case_numbers <- "24,29,32,34,35,42,47,52,54,59,72,79,88,95,98,99,100"
control_numbers <-"49114,49109,49106,49104,49103,49096,49092,49088,49086,49081,49069,49062,49053,49046,49043,49042,49041"

n_tb <- bind_cols('lagtime'=seq(0,16),as.numeric(str_split_1(case_numbers,',')), as.numeric(str_split_1(control_numbers,',')))
n_tb <- mutate(n_tb,lagtime = as_factor(lagtime))
colnames(n_tb) <- c('lagtime', 'case_n','control_n')

#Plot of case number over lagtime
plot_output_path <- 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/'


n_plotter <- function(input_tb, plot_output_path, column_name='case_n'){
  
  if (column_name=='case_n'){label <- 'Cases'} else if (column_name=='control_n'){label <- 'Controls'}
  
  max_n = max(input_tb[[column_name]])
  
  n_plot <- ggplot(input_tb,aes(x=lagtime, y=.data[[column_name]]))+
    geom_col()+
    xlab('Lag Time Assumed (Years)')+
    ylab(str_c('Number of ', label))+
    geom_text(aes(label=.data[[column_name]]), nudge_y=max_n/20) #Dynamic height adjustment to 1/20 of the max value
  
  print(n_plot)
  ggsave(str_c(plot_output_path, label,'_n.png'), dpi=600, width=9, height = 3)
  
}

n_plotter(n_tb, plot_output_path)
# n_plotter(n_tb, plot_output_path, 'control_n')
```


# Case-Control Comparison of Different Lagtimes 

```{r import_limma}
limma_importer <- function(filedir){
  f <- list.files(filedir, full.names=F, pattern='limma_results')
  lagtimes <- as.numeric(str_match(f, '(\\d{1,2})Y_lag')[,2])
  
  results <- map2(str_c(filedir,f), lagtimes, ~read_tsv(.x) %>% mutate(lag=.y)) %>% bind_rows()
  return(results)
}

lag_results <- limma_importer('OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/')
```

## Adjust for Inflation in Test Statistics

```{r}
#This checks for the lambda inflation factor for each of the lag times tested
lambda_checker <- function(results_df){

  # This quantifies the deviation you see in the Q-Q plot.
  chisq_stats <- qchisq(results_df$P.Value, df = 1, lower.tail = FALSE) #This step recomputes the Chi-sq test statistic from the p-value itself assuming the p-value follows a Chi-sq distribution with df=1
  lambda <- median(chisq_stats) / qchisq(0.5, df = 1)

  return(lambda)
}

lambdas <- map(lagtimes, ~lambda_checker(filter(lag_results, lag == .)))
names(lambdas) <- str_c(lagtimes, ' Y')

print(sum(lambdas > 1.1, na.rm=T)) #How many lag times have lambda > 1.1
```

```{r}
inflation_adjuster <- function(limma_results, lambda){

  #This readjusts the p-values for inflation adjustment by computing the Chi-squared test statistic from the p-value and then dividing by lambda -> recompute the p-values
  limma_results <- limma_results %>%
    mutate(lambda_adjusted_pvalue = pchisq(qchisq(limma_results$P.Value, df = 1, lower.tail = FALSE)/lambda,df=1,lower.tail=FALSE))
  
  return(limma_results)
}

lag_results_inflationadj <- map2(lagtimes, lambdas, ~inflation_adjuster(filter(lag_results, lag == .x), .y))

#Convert list back to a single tibble
lag_results <- lag_results_inflationadj %>% bind_rows()
```

Extract out all plasma proteins that are 5% FDR-significant in at least one time lag instance.

```{r}
lag_results <- lag_results %>%
  group_by(lag) %>%
  mutate(fdr_pval = p.adjust(lambda_adjusted_pvalue, method='BH')) %>%
  ungroup() %>%
  mutate(fdr5_pass_all_lags = ifelse(fdr_pval <= 0.05, T, F)) %>%
  group_by(pp) %>%
  mutate(num_instances_sig = sum(fdr_pval <= 0.05)) %>%
  ungroup() %>%
  mutate(pp = reorder(factor(pp,ordered=T), num_instances_sig)) %>%
  mutate(ln_abs_beta = log1p(abs(B))) %>%
  mutate(neglog10_fdr_pval = -log10(fdr_pval))

fdr5_proteins_all_lag <- unique(filter(lag_results, fdr5_pass_all_lags==T)$pp) #5 plasma proteins are FDR5-sig. in at least one time instance
```

Output a plot showing how the betas and FDR-adjusted(-log10) p-values change as a function of the time lag.

```{r}
timelag_plotter <- function(input_tb, metric, metric_name, pp_of_interest, output_folder, fc_ylims = c(0,2), suffix=''){
  
  plot_tb <- input_tb %>%
    filter(pp %in% pp_of_interest)
  
  #Heatmap
  heatmap_plot <- ggplot(plot_tb, aes(x=lag, y=pp, fill=!!sym(metric)))+
    geom_tile()+
    viridis::scale_fill_viridis(direction=ifelse(metric=='fdr_pval', -1, 1))+
    xlab('Lag Time Assumed (Years)')+
    ylab('Plasma Protein')+
    labs(title = str_wrap(str_c('Heatmap of plasma proteins passing 5% FDR in at least one instance over ', length(unique(plot_tb$lag)), 'lag time instances with colour = ', metric_name)))+
    labs(fill=metric_name)

  print(heatmap_plot)
  ggsave(str_c(output_folder, 'summary_',metric, '_heatmap',suffix,'.png'), heatmap_plot, dpi=600, width = 6, height = 6)
  
  #Line plot - facetted by the number of time instances in which the protein is FDR-significant
  #Add labels to each line at the last instance
  max_instance <- max(plot_tb$lag)
  plot_tb <- plot_tb %>%
    mutate(Label = ifelse(lag==max_instance, as.character(pp), ''))
  
    # Get unique facet levels
  n_sig_levels <- unique(plot_tb$num_instances_sig)
  
  # Base plot
  p <- ggplot()
  
  # Loop over each facet level and add a new layer
  for (n in n_sig_levels) {
    p <- p +
      geom_step(data = plot_tb %>% filter(num_instances_sig == n), 
                aes(x = lag, y = !!sym(metric), color = pp)) +
      geom_text_repel(data = plot_tb %>% filter(num_instances_sig == n), 
                aes(x = lag, y = !!sym(metric), color = pp, label=Label),
                force=3, max.overlaps=100)+
      scale_color_brewer(palette = "Paired") +
      scale_x_continuous(breaks = seq(from = min(plot_tb$lag), to = max(plot_tb$lag), by = 2))+
      xlab('Lag Time Assumed (Years)')+
      ylab(metric_name)+
      theme(legend.position = "none")+
      # guides(color = guide_legend(title = "Plasma Protein", direction = 'horizontal', position='bottom'))+
      new_scale_color()  # Reset colors for the next facet
    
    if(metric == 'logFC'){
      p <- p+
        scale_y_continuous(limits=fc_ylims)
    }

    if(metric=='fdr_pval'){
      p <- p+
        geom_hline(yintercept=0.05,linestyle='dashed', alpha=0.25, col='red', linetype='dashed')
    } else if (metric=='neglog10_fdr_pval'){
      p<-p+
        geom_hline(yintercept=-log10(0.05),linestyle='dashed', alpha=0.25, col='red', linetype='dashed')
  }
  
  # Add facets and theme
  p <- p + facet_wrap(~num_instances_sig, scales = "fixed", ncol=4)+
        labs(title=str_c("Line plot of plasma proteins' " , metric_name, ' over ', length(unique(plot_tb$lag)),' lag time instances facetted by number of lag times in which it is 5% FDR-significant'))
    }
  
  print(p)
  
  ggsave(str_c(output_folder, 'summary_',metric, '_lineplot',suffix,'.png'), p, dpi=600, width = 12, height = 5)
  
}

timelag_plotter(lag_results, 'logFC', 'log2(Fold-Change)',fdr5_proteins_all_lag, 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/')
# timelag_plotter(lag_results, 'ln_abs_beta', 'Ln(|Beta|)',fdr5_proteins_all_lag, 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/')
# timelag_plotter(lag_results, 'fdr_pval', 'FDR-adjusted p-value',fdr5_proteins_all_lag, 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/')
timelag_plotter(lag_results, 'neglog10_fdr_pval', '-log10(FDR-adjusted p-value)',fdr5_proteins_all_lag, 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/')

#Ignore NTproBNP and NPPB
# timelag_plotter(lag_results, 'logFC', 'log2(Fold-Change)',fdr5_proteins_all_lag[!fdr5_proteins_all_lag %in% c('ntprobnp','nppb')], 'OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/0_TimeInstancing/HCM/rcmm_cov/allpanels/', suffix='_wo_NTproBNP_NPPB')
```

# Incident Cox HCM Diagnosis

```{r}
multivariate_cox <- function(input_tb, marginal_predictor, covariates, status, outcome_var, outcome_levels, print_summary=T,return_predictor_results_only=T, return_model=F){ #Return only the coefficient and p-values for the marginal_predictor
  
  x <- str_c(c(covariates,marginal_predictor), collapse='+')
  y <- str_c('Surv(',outcome_var,',',status,')')
  formula <- as.formula(paste(y,'~',x))
  
  multivariate_cox <- coxph(formula, data=input_tb)
  if(isTRUE(print_summary)){
    print(summary(multivariate_cox))
  }
  
  x<- summary(multivariate_cox)
  
  if(isTRUE(return_model)){
    return(multivariate_cox)
  }
  
  if(isFALSE(return_predictor_results_only)){
  
  p.value<-signif(x$coef[1:nrow(x$coef),5], digits=3)
  beta<-signif(x$coef[1:nrow(x$coef),1], digits=3);#coeficient beta
  HR <-signif(x$coef[1:nrow(x$coef),2], digits=3);#exp(beta) = hazard ratio
  HR.confint.lower <- signif(x$conf.int[,"lower .95"], 3)
  HR.confint.upper <- signif(x$conf.int[,"upper .95"],3)
  se_beta <- signif(x$coef[1:nrow(x$coef),3], digits=3)
  vars <- rownames(x$coef)
    
  } else{
      p.value<-signif(x$coef[nrow(x$coef),5], digits=3)
  beta<-signif(x$coef[nrow(x$coef),1], digits=3);#coeficient beta
  HR <-signif(x$coef[nrow(x$coef),2], digits=3);#exp(beta) = hazard ratio
  HR.confint.lower <- signif(x$conf.int[,"lower .95"][nrow(x$coef)], 3)
  HR.confint.upper <- signif(x$conf.int[,"upper .95"][nrow(x$coef)],3)
  se_beta <- signif(x$coef[nrow(x$coef),3], digits=3)
  vars <- rownames(x$coef)[nrow(x$coef)]

  }
  
  res<-bind_cols(vars,beta, se_beta, HR, HR.confint.lower, HR.confint.upper, p.value) %>%
    mutate(predictor=str_to_upper(vars)) %>%
    mutate(outcome = factor(
      str_to_upper(str_match(outcome_var,'(.+)_age')[,2]),levels=outcome_levels, ordered=T
    ))
  
  names(res)<-c('var',"beta",'se_beta', "HR", "Lower_95CI_HR","Upper_95CI_HR", 
            "p.value", 'predictor','outcome')
  
  return(res)
}

multivariate_coxph_plotter <- function(multivariate_results, plot_output_path, output_name, pp_scaling, FDR=5, specific_label=F, ylimits=''){
  
  if (!is.numeric(FDR)){ #Apply Bonferroni
    pval_threshold = 0.05/nrow(multivariate_results)
    mtc <- 'Bonferroni'
  } else{
    multivariate_results <- mutate(multivariate_results, p.value=p.adjust(p.value, method='BH'))
    pval_threshold <- FDR/100
    mtc <- str_c('FDR ',FDR, '%',sep='')
  }
  
  if(isFALSE(specific_label)){
    multivariate_results <- multivariate_results %>% #Only label the ones which reach significance
      mutate(Label=ifelse(p.value < pval_threshold,Label, ''),
           MTC_significant = ifelse(p.value < pval_threshold, T, F)) %>%
      arrange(Label)
  } else {
    multivariate_results <- multivariate_results %>%
      mutate(MTC_significant = ifelse(Label=='', F, T)) %>%#Just for downstream code purposes to label some specific points
      arrange(Label)
  }
  
  multivariate_plot <- ggplot(multivariate_results, aes(HR,-log10(p.value), col=MTC_significant))+
    geom_vline(xintercept=1, linetype='dashed',alpha=0.5)+
    geom_hline(yintercept=-log10(pval_threshold), linetype='dashed', alpha=0.5)+
    geom_point(alpha=0.75)+
    scale_colour_manual(name=ifelse(isFALSE(specific_label),'MTC Significant','PP of Interest'),values=c('FALSE'='grey','TRUE'='red'))+
    geom_text_repel(aes(label=Label), nudge_y=3, force = 5, max.overlaps=40)+
    #geom_errorbarh(aes(xmin=Lower_95CI_HR, xmax=Upper_95CI_HR,col=predictor),alpha=0.75, height=0.1)+
    xlab(ifelse(isFALSE(pp_scaling),'Hazard Ratio', 'Hazard Ratio per SD'))+
    ylab(ifelse(mtc=='Bonferroni','-log10 (Wald-test p value)', '-log10 (Adjusted Wald-test p value)'))+
    scale_x_continuous(n.breaks=10)+
    labs(title=str_wrap(str_c('Multivariate Cox Regression analysis for ', nrow(multivariate_results), ' plasma protein predictors')),caption=str_c(mtc, ' p-value Threshold Indicated'))+
    theme(legend.position="right")
  
  if(!identical(ylimits,'')){
    multivariate_plot <- multivariate_plot+
      scale_y_continuous(limits=ylimits)
  }
  
  print(multivariate_plot)
  ggsave(str_c(plot_output_path, 'cox_multivariate_', output_name,'.png'), dpi=600, width=9, height=6)
  
  return(multivariate_results)
}
```

## Main2

```{r}
ukb_outcomes <- read_tsv('../../HCM_Severity/DATASETS/UKB/RAP/Outcomes/Processed/allukb_outcome_ages.tsv')
ukb_cases_total <- read_tsv('DATA/UKB/RAP/Cases/ukb_rap_HCM_20240209.csv')

main2 <- function(lagtime,  #Lagtime to test in years
                  ukb_pp_list, #PPs to include in analysis,
                  ukb_outcomes,
                  hcm_diag_tb, #This contains the minimum diagnosis dates
                  ukb_cases_total,
                  ukb_basic_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_age_sex_htwt_10pcs_dbp_t2d_smoking_cov.tsv',  
                  ukb_rcmm_cov_filepath='DATA/UKB/PROCESSED/20240209_ukb_rcmm_cov.tsv',
                  ukb_pp_filepath='DATA/UKB/RAP/PP/ukb_allpp_i0.tsv', 
                  pp_scaling='ranknorm',
                  rcmm_cov=T,
                  include_nonCM_panels=T,
                  output_basepath = 'OUTPUT/UKB/PLOTS/5_CoxHCMDiag/0_TimeInstancing/'){
  
  hcm_cases <- hcm_case_instance_calculator(hcm_diag_tb, 'HCM', lagtime=lagtime)
  
  ukb_pp_cov_hcm <- overall_tb_builder(hcm_cases, #HCM cases in UKB from the particular instance
                                str_c(lagtime, 'Y', collapse = ''), #Name of the instance
                                ukb_pp_list, #List of plasma proteins to compare
                                ukb_basic_cov_filepath,  
                                ukb_rcmm_cov_filepath,
                                ukb_pp_filepath, 
                                pp_scaling,include_nonCM_panels, rcmm_cov)
  
    pp <- ukb_pp_list$ukb_fieldname_equivalent[ukb_pp_list$ukb_fieldname_equivalent %in% colnames(ukb_pp_cov_hcm)] %>%  #In case any PP are removed due to lack of non-NA values
  unique()
  
  #Make further changes to the ukb_pp_cov_hcm to prepare it for CoxHCMdiag
    if (pp_scaling == 'scale'){
    pp_tb <- ukb_pp_cov_hcm %>% select(all_of(pp)) %>%
      scale()
    
    ukb_pp_cov_hcm <- ukb_pp_cov_hcm %>%
      select(-all_of(pp)) %>%
      bind_cols(pp_tb)
      
    rm(pp_tb)
    
  }else if (pp_scaling == 'ranknorm'){
    ukb_pp_cov_hcm <- ukb_pp_cov_hcm %>%
      mutate(across(all_of(pp), ~ replace(., !is.na(.), RNOmni::RankNorm(.[!is.na(.)]))))
  }
  
  ukb_diag_cox_tb <- ukb_pp_cov_hcm %>%
  left_join(select(ukb_outcomes,ID, lost_age, death_age, age_attend_i0, dob_approx), by=c('eid'='ID')) %>%
  left_join(select(hcm_diag_tb, ID,min_diag_date2), by=c('eid'='ID'))

  ukb_diag_cox_tb <- ukb_diag_cox_tb %>%
    filter(!eid %in% hcm_cases$ID) %>% #Filter out HCM cases who were diagnosed prior to the blood sample collection date + lag time assumed
    mutate(datasetversion_age = time_length(ymd('2024-02-09') - dob_approx, unit='year')) %>%
    mutate(incidenthcm_status = ifelse (eid %in% ukb_cases_total[['0']], T,F)) %>% #Add in status
    rowwise() %>%
    mutate(incidenthcm_age = (ifelse(incidenthcm_status==T, time_length(min_diag_date2-dob_approx, unit='year'), min(datasetversion_age, death_age, lost_age, na.rm=T)))) %>% #Add in incident_hcm_age = if status, that will be min_diag_date2 and if not, then it will be the age at dataset version or lost to followup
    ungroup()
  
  if (isTRUE(rcmm_cov)){  #If using the extended covariate list suggested by Rob Clarke and Mohsen
    extra_folders <- 'rcmm_cov/'
    covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking','Townsend','eGFR', 'ldl')
  } else {
    covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking')
  }
  if(isTRUE(include_nonCM_panels)){ extra_folders <- str_c(extra_folders, 'allpanels/')} 
  
  path <- str_c(output_basepath, extra_folders) #Define the folder for output of plots and tabular data
  if(!dir.exists(path)){dir.create(path, recursive=T)}
  
    
  multivariate_cox_results <- c(
    map(pp, ~multivariate_cox(ukb_diag_cox_tb,.,covars,'incidenthcm_status','incidenthcm_age', outcome_levels=c('INCIDENTHCM'), print_summary = F))
  ) %>%
    bind_rows() %>%
    filter(!is.na(p.value))
  
  multivariate_cox_results <- multivariate_cox_results %>%
    left_join(dplyr::select(ukb_pp_list,ukb_fieldname_equivalent,Target), by=c('var'='ukb_fieldname_equivalent')) %>%
    unique() %>%
    dplyr::rename(Label=Target)
  
  saveRDS(multivariate_cox_results, str_c(path,'/mvCox_results_pp',pp_scaling,'_',lagtime, 'Y.rds'))
  
  mvcox_incidentHCM_fdr5 <- multivariate_coxph_plotter(multivariate_cox_results, path, str_c('pp_incidentHCMdiag_fdr5_pp',pp_scaling,'_',lagtime,'Y'), pp_scaling=pp_scaling)
  multivariate_coxph_plotter(multivariate_cox_results, path, str_c('pp_incidentHCMdiag_fdr1_pp',pp_scaling,'_',lagtime,'Y'), FDR=1, pp_scaling=pp_scaling)
  multivariate_coxph_plotter(multivariate_cox_results, path, str_c('pp_incidentHCMdiag_bonferroni_pp', pp_scaling,'_',lagtime,'Y'), FDR=F, pp_scaling=pp_scaling)
  
  #Write out 
  write_tsv(dplyr::select(filter(ukb_pp_list, ukb_fieldname_equivalent %in% filter(mvcox_incidentHCM_fdr5, MTC_significant==T)$var),UniProt),
            str_c(path,'fdr5_pass_uniprotIDs_pp',pp_scaling,'_',lagtime,'Y.tsv'), col_names=F)
  
}

lagtimes <- c(2,3,4,seq(6,17))
walk(lagtimes, ~main2(., ukb_pp_list, ukb_outcomes, hcm_diag_tb, ukb_cases_total))

```

# Incident Comparison
```{r}
diagrds_importer <- function(filedir){
  f <- list.files(filedir, full.names=F, pattern='mvCox_results')
  lagtimes <- as.numeric(str_match(f, '_(\\d{1,2})Y.rds')[,2])
  
  results <- map2(str_c(filedir,f), lagtimes, ~readRDS(.x) %>% mutate(lag=.y)) %>% bind_rows()
  return(results)
}

cox_lag_results <- diagrds_importer('OUTPUT/UKB/PLOTS/5_CoxHCMDiag//0_TimeInstancing/rcmm_cov/allpanels/')

#Filter out outlier year of 14Y time instance (some instability in Cox regression likely)
cox_lag_results <- cox_lag_results %>%
  filter(lag <= 12)

cox_lag_results <- cox_lag_results %>%
  # dplyr::rename(pp=var) %>%
  group_by(lag) %>%
  mutate(fdr_pval = p.adjust(p.value, method='BH')) %>%
  ungroup() %>%
  mutate(fdr5_pass_all_lags = ifelse(fdr_pval <= 0.05, T, F)) %>%
  group_by(pp) %>%
  mutate(num_instances_sig = sum(fdr_pval <= 0.05)) %>%
  ungroup() %>%
  mutate(pp = reorder(factor(pp,ordered=T), num_instances_sig)) %>%
  mutate(ln_abs_beta = log1p(abs(beta))) %>%
  mutate(neglog10_fdr_pval = -log10(fdr_pval))

fdr5_Cox_pp_all_lag <- as.character(unique(filter(cox_lag_results, fdr5_pass_all_lags==T)$pp)) #11 plasma proteins are FDR5-sig. in at least one time instance
# fdr5_Cox_pp_all_lag_thresh2 <- as.character(unique(filter(cox_lag_results, num_instances_sig >=2)$pp)) #7 plasma proteins pass FDR significance in at least 2 time instances
```

```{r}
timelag_plotter(cox_lag_results, 'beta', 'Beta',fdr5_Cox_pp_all_lag, 'OUTPUT/UKB/PLOTS/5_CoxHCMDiag/0_TimeInstancing/rcmm_cov/allpanels/')
timelag_plotter(cox_lag_results, 'HR', 'Hazard Ratio',fdr5_Cox_pp_all_lag, 'OUTPUT/UKB/PLOTS/5_CoxHCMDiag/0_TimeInstancing/rcmm_cov/allpanels/')
```


